---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)

library(ggplot2)            # all tasks
library(scales)             # needed for task 2
library(RColorBrewer)       # needed for task 2
library(ggrepel)            # trying something out in task 3
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

**Solution:**

Load the dataset `expeditions` from the provided online source:

```{r}
expeditions <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/expeditions.csv")
```

Take a look at the dataset `expeditions`.

```{r}
glimpse(expeditions)
```

OK, let's continue with the actual plotting! **My logic for the task:**

-   I first thought of just `count()` with sorting, and `slice(1:15)`. But that would require more steps, namely extracting 15 most frequent names in a separate vector, then filtering the main dataset by filter(name %in% top_15). Then I followed the instructions that hint to use `fct_lump()` = we lump 15 most frequent `peak_names` and lump others to `Other` and then `filter()` it out as we won't need other peaks.
-   I planned to use `group_by(peak)`, count number of rows with `n()` and `ungroup()` before next steps, but after looking at the documentation online, I came across wonderful function `add_count()`! Then, it was easy to just use `fct_reorder()` on the whole/remaining dataset!
-   As per example plot, it was straightforward to use `geom_bar()` with `season` as fill argument. Coordinates needed to be flipped, labels added, legend re-positioned and after looking at the online sources, I figured how to add the required color palette (viridis).

```{r}
expeditions %>%
  mutate(peak = fct_lump(peak_name, 15)) %>%
  filter(peak != "Other") %>%
  #we could group_by peak and use n() to calculate total rows per peak, but this is neater
  add_count(peak) %>%    
  mutate(peak = fct_reorder(peak, n)) %>%
  ggplot() + 
  aes(peak, fill = season) +
  geom_bar() + 
  coord_flip() +
  theme_light() + 
  labs(x = NULL, y = "Number of expeditions", title = "The 15 most popular peaks by season of expedition") + 
  theme(legend.position="bottom") + 
     scale_fill_viridis_d()
  
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

**Solution:**

Import the dataset from the provided source:

```{r}
phd_field <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")
```

Take a look at the dataset `phd_field`:

```{r}
glimpse(phd_field)
```

OK, let's continue with the actual plotting! **My logic for the task:**

-   Based on the instructions, I know that we need `broad_field`, `major_field` and `field` as factors. So, let's factorise and overwrite existing variables.
-   *What data do I need?* Sum of all phds (`n_phds`) by `year` by `broad_field` (first 2 lines)
-   *This summarised data is sufficient for the plot.* I need `year`on X axis, summarised `total_phds` on Y axis, geom_line() requires grouping aesthetic (`broad_field`) and we also assign the same (`broad_field`) to color aesthetic to get lines of different colors
-   Search online how to assign `pretty_breaks()`, `comma_format()` and `scale_color_brewer()`
-   Change the labels according to the provided example

```{r}
phd_field %>%
  mutate(broad_field = factor(broad_field),
         major_field = factor(major_field),
         field = factor(field)) %>%
  group_by(broad_field, year) %>%
  summarise(total_phd = sum(n_phds, na.rm = TRUE)) %>%
  ggplot() +
  aes(x = year, y = total_phd, group = broad_field, color = broad_field) +
  geom_line(size = 1.2) + 
  theme_minimal() + 
  scale_x_continuous(breaks = pretty_breaks()) +    
  scale_y_continuous(labels = comma_format()) + 
  scale_color_brewer(palette = "Dark2") + 
  labs(x = NULL, y = NULL, color = "Broad field", title = "Number of awarded Ph.D.-s in the US by year")
  
```


## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

**Solution:**

Compared to previous 2 tasks, I divided this one into more steps. Could be done in one pipeline, but it might be a bit messy to read/follow. Edit: For practice I did everything in one step in the end, by far not as *bad* as I expected....actually very easy to follow :)


Read the dataset from the online source:
```{r}
commute <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2019/2019-11-05/commute.csv")
```

Take a look at the data:
```{r}
glimpse(commute)
```

Factorise specific variables:
```{r}
commute <- commute %>%
  mutate(city = factor(city),
         state = factor(state),
         city_size = factor(city_size),
         mode = factor(mode),
         state_abb = factor(state_abb),
         state_region = factor(state_region))
```

*What data do I need?* `state_abb` for point labeling, `state_region` for grouping (`fill`), sum of `Walk` and `Bike` for each state. 
```{r}
commute_sum <- commute %>%
  group_by(state_region, state_abb, mode) %>%
  summarise(total_number = sum(n))
```

Ok, getting there. But, we need to plot `Bike` vs. `Walk`, so we will need to `pivot_wider()` (good that we practiced pivoting in the previous assignment a bit).
```{r}
commute_wide <- pivot_wider(commute_sum, names_from = mode, values_from = total_number)
```

Now we can finally plot!
```{r}
commute_wide %>%
  ggplot() + 
  aes(x = Walk, y = Bike, color = state_region) + 
  geom_point(size = 2) + 
  geom_text_repel(aes(label = state_abb), color = "black", size = 3) + 
  scale_x_log10(labels = label_comma()) + 
  scale_y_log10(limits = c(80, 150000),labels = label_comma()) +
  theme_light() + 
  labs(x = "Number of ppl walking walking to work (log10)", y = "Number of ppl walking biking to work (log10)", color = "State region", title = "Number of people walking vs. biking to work in each USA state")
```


Ok for practice I want to see how all these steps would look in one pipeline:
```{r}
commute %>%
  mutate(mode = factor(mode),
         state_abb = factor(state_abb),
         state_region = factor(state_region)) %>%
  group_by(state_region, state_abb, mode) %>%
  summarise(total_number = sum(n)) %>%
  pivot_wider(names_from = mode, values_from = total_number) %>%
  ggplot() + 
  aes(x = Walk, y = Bike, color = state_region) + 
  geom_point(size = 2) + 
  geom_text_repel(aes(label = state_abb), color = "black", size = 3) + 
  scale_x_log10(labels = label_comma()) + 
  scale_y_log10(limits = c(80, 150000),labels = label_comma()) +
  theme_light() + 
  labs(x = "Number of ppl walking walking to work (log10)", y = "Number of ppl walking biking to work (log10)", color = "State region", title = "Number of people walking vs. biking to work in each USA state")
  
```
Ok....not **that** bad actually, not bad at all.
